---
title: "FFCM_2025_Browse"
output:
  word_document: default
  html_document: default
date: "2025-02-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, echo = FALSE, warning=FALSE, message=FALSE}
# Loading in packages
#install.packages("readxl")
#install.packages('tidyverse')
#install.packages('dplyr')
#install.packages('writexl')
#install.packages('ggplots')
#install.packages('forcats')
library(readxl)
library(tidyverse)
library(dplyr)
library(writexl)
library(ggplot2)
library(forcats)
```

```{r master wide, echo=FALSE, warning = FALSE}
master_wide <- read_excel("C:\\Users\\jattanasio\\Desktop\\R_related\\master_wide.xlsx")
```

Note: Belfast does not need to be separated because I am only looking at 2024 browse.

```{r, echo=FALSE, warning = FALSE}
browse <- master_wide %>%
  select(Browse_fall2024, LiveDead_fall2024, Region, UniqueID, Site, Plot, Species)  %>%
  filter(LiveDead_fall2024 == "L") %>%
  mutate(Browse_fall2024 = replace_na(Browse_fall2024, "N"))

ggplot(browse, aes(x = Species)) +
  geom_bar(aes(fill = Browse_fall2024)) +
   scale_fill_manual(values = c("#66B032", "#B25600")) +
  facet_wrap(~Site) +
  labs(title = "Browse in 2024", x = "Species", y = "Number of Seedlings", fill = "Browse") +
  theme(axis.text.x = element_text(angle = 90))
```

Interesting how overall Surry does not have a lot of browse compared to Schoodic
that has higher proportion of browse. Also obvious in MDI how the broadleaf seedlings
are getting browsed a lot more than the needles.

It seems like Schoodic has had more seedlings die than other sites, because
there seem to have fewer seedlings compared to other sites. Will look into this more
when I look at live vs dead.

Also important to keep in mind that Schoodic has a lot of browse, and it is probably due
to hare and not deer. I wonder if hare browse different species than deer.

From experience I know deer prefer the deciduous seedlings: oaks, tulip, sweet gum

Apparently Snowshoe hares (which is what we have in the area) feed on deciduous trees and shrubs like birch, willow, maple. Also some conifers like white pine and balsam fir (according to Canadian Fish and Wildlife). It would also make sense if they eat spruce

```{r, echo=FALSE, warning = FALSE, message=FALSE}
# How many of each species was browsed (in fall2024 and were alive then)
species_browse <- browse %>%
  filter(LiveDead_fall2024 == "L") %>%
  group_by(Species, Browse_fall2024) %>%
  summarise(n = n())
# percent of browse per site
site_browse <- browse %>%
  filter(LiveDead_fall2024 == "L") %>%
  group_by(Site, Browse_fall2024) %>%
  summarise(n = n())
```


## Percent Browse Per Species in 2024

|Species |Region  |Planted |Live    |Browse  |
|--------|--------|--------|--------|--------|
|s.gum   |southern|193     |59      |45.8%   |
|tulip   |southern|192     |49      |36.7%   |
|r.cedar |southern|180     |47      |0%      |
|ch.oak  |southern|192     |118     |49.2%   |
|w.oak   |southern|186     |87      |43.7%   |
|r.oak   |northern|192     |55      |38.2%   |
|w.pine  |northern|191     |110     |4.5%    |
|w.spruce|northern|192     |98      |4.1%    |

## Percent Browse Per Site in 2024

|Site    |Planted |Live    |Browse  |
|--------|--------|--------|--------|
|Belfast |380     |227     |23.8%   |    
|MDI     |383     |150     |36.4%   |  
|Schoodic|374     |68      |69.1%   |
|Surry   |381     |177     |8.5%    |

> Overall take-aways from browse

> - Red cedar was not browsed at all, and white pine and white spruce have very low browse
> - Species with most browse was chestnut oak (has the highest survival though)
> - 69.1% of seedlings browsed in Schoodic 2024 (most browse)
> - 8.5% of seedlings browsed in Surry 2024 (least browse)






